Visual Diagnostics for Constrained Optimisation with Application to Guided Tours

### 2021-03-08 paper/zhang-cook-laa-langrene-menendez.Rmd
more changes

Let $\mathbf{X}_{n \times p}$ be the data matrix with $n$ observations in $p$ dimensions. A $d$-dimensional projection is a linear transformation from $\mathbb{R}^p$ into $\mathbb{R}^d$ defined as $\mathbf{Y} = \mathbf{X} \cdot \mathbf{A}$, where $\mathbf{Y}_{n \times d}$ is the projected data and $\mathbf{A}_{p\times d}$ is the projection basis matrix. We define $f: \mathbb{R}^{n \times d} \mapsto \mathbb{R}$ to be an index function that maps the projected data $\mathbf{Y}$ onto a scalar value. This is commonly known as the projection pursuit index function, or just index function, and is used to measure the "interestingness" of a given projection. In projection pursuit, higher index values are associated with interesting projections that depart from normality [@diaconis1984asymptotics; @huber1985projection], e.g. Legendre index [@friedman1974projection], Hermite index [@hall1989polynomial] and natural Hermite index [@cook1993projection] for measuring the distance between the density of projected data and a standard normal distribution, LDA index [@lee2005projection] and PDA index [@lee2010projection] for finding clustering structure in the data, and skewness-based index [@Loperfido2018] and kurtosis-based index [@Loperfido2020] for outlier detection. As a general visualisation method, tour produces animations of high dimensional data via rotations of low-dimensional planes. There are different tour implementations depending on how the high dimensional space is investigated: grand tour [@cook2008grand] selects the planes randomly to provide a general overview;  manual tour [@cook1997manual] gradually phases in and out one variable to understand the contribution of that variable in the projection. Guided tour, the main interest of this paper, chooses the planes with the aid of projection pursuit, to gradually reveal the most interesting projection. Given a random start, projection pursuit iteratively finds bases with higher index values and the guided tour constructs a geodesic interpolation between these planes to form a tour path. Figure \ref{fig:tour-path} shows a sketch of the tour path where the blue frames are produced by the projection pursuit optimisation and the white frames are interpolations  between the blue frames. Mathematical details of the geodesic interpolation can be found in @buja2005computational. The aforementioned tour method has been implemented in the R package \CRANpkg{tourr} [@tourr].

@@

\noindent where $f$, $\mathcal{A}$, and $\mathbf{X}$ are defined as in the previous section,  $I_d$ is the $d$-dimensional identity matrix and the constraint ensures the projection bases $\mathbf{A}$ to be orthonormal. It is worth noticing the following: 1) The optimisation is constrained and the basis space is a $d$-dimensional sphere. 2) There may be index functions for which the objective function might not be differentiable. 3) While finding the global optimum is the goal of the optimisation problem, interesting projections may also appear in the local optimum. 4) The optimisation should be computationally agile since the tour animation is viewed by the users during the optimisation.

>>>>

Let $\mathbf{X}_{n \times p}$ be the data matrix with $n$ observations in $p$ dimensions. A $d$-dimensional projection is a linear transformation from $\mathbb{R}^p$ into $\mathbb{R}^d$ defined as $\mathbf{Y} = \mathbf{X} \cdot \mathbf{A}$, where $\mathbf{Y}_{n \times d}$ is the projected data and $\mathbf{A}_{p\times d}$ is the projection basis matrix. We define $f: \mathbb{R}^{n \times d} \mapsto \mathbb{R}$ to be an index function that maps the projected data $\mathbf{Y}$ onto a scalar value. This is commonly known as the projection pursuit index function, or just index function, and is used to measure the "interestingness" of a given projection. In projection pursuit, higher index values are associated with interesting projections that depart from normality [@diaconis1984asymptotics; @huber1985projection], e.g. Legendre index [@friedman1974projection], Hermite index [@hall1989polynomial] and natural Hermite index [@cook1993projection] for measuring the distance between the density of projected data and a standard normal distribution, LDA index [@lee2005projection] and PDA index [@lee2010projection] for finding clustering structure in the data, and skewness-based index [@Loperfido2018] and kurtosis-based index [@Loperfido2020] for outlier detection.

As a general visualisation method, tour produces animations of high dimensional data via rotations of low-dimensional planes. There are different tour implementations depending on how the high dimensional space is investigated: grand tour [@cook2008grand] selects the planes randomly to provide a general overview;  manual tour [@cook1997manual] gradually phases in and out one variable to understand the contribution of that variable in the projection. Guided tour, the main interest of this paper, chooses the planes with the aid of projection pursuit, to gradually reveal the most interesting projection. Given a random start, projection pursuit iteratively finds bases with higher index values and the guided tour constructs a geodesic interpolation between these planes to form a tour path. Figure \ref{fig:tour-path} shows a sketch of the tour path where the blue frames are produced by the projection pursuit optimisation and the white frames are interpolations  between the blue frames. Mathematical details of the geodesic interpolation can be found in @buja2005computational. The aforementioned tour method has been implemented in the R package \CRANpkg{tourr} [@tourr].

@@

\noindent where $f$ and $\mathbf{X}$ are defined as in the previous section, $\mathcal{A}$ is the set of all $p$-dimensional projection bases, $I_d$ is the $d$-dimensional identity matrix, and the constraint ensures the projection bases, $\mathbf{A}$, to be orthonormal. It is worth noticing the following: 1) The optimisation is constrained and the orthonormality constraint imposes a geometrical structure on the bases space: it forms a Stiefel manifold. 2) There may be index functions for which the objective function might not be differentiable. 3) While finding the global optimum is the goal of the optimisation problem, interesting projections may also appear in the local optimum. 4) The optimisation should be computationally agile since the tour animation is viewed by the users during the optimisation.

### 2021-03-08 paper/zhang-cook-laa-langrene-menendez.tex
more changes

\noindent where \(f\), \(\mathcal{A}\), and \(\mathbf{X}\) are defined
as in the previous section, \(I_d\) is the \(d\)-dimensional identity
matrix and the constraint ensures the projection bases \(\mathbf{A}\) to
be orthonormal. It is worth noticing the following: 1) The optimisation
is constrained and the basis space is a \(d\)-dimensional sphere. 2)
There may be index functions for which the objective function might not
be differentiable. 3) While finding the global optimum is the goal of
the optimisation problem, interesting projections may also appear in the
local optimum. 4) The optimisation should be computationally agile since
the tour animation is viewed by the users during the optimisation.

>>>>

\noindent where \(f\) and \(\mathbf{X}\) are defined as in the previous
section, \(\mathcal{A}\) is the set of all \(p\)-dimensional projection
bases, \(I_d\) is the \(d\)-dimensional identity matrix, and the
constraint ensures the projection bases, \(\mathbf{A}\), to be
orthonormal. It is worth noticing the following: 1) The optimisation is
constrained and the orthonormality constraint imposes a geometrical
structure on the bases space: it forms a Stiefel manifold. 2) There may
be index functions for which the objective function might not be
differentiable. 3) While finding the global optimum is the goal of the
optimisation problem, interesting projections may also appear in the
local optimum. 4) The optimisation should be computationally agile since
the tour animation is viewed by the users during the optimisation.
